Learning words through recursive pragmatic reasoning about other agents

### 2013-05-31 praglearn.tex
modifications

\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}

@@

\section{Experiments 1: Empirical coverage}

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\hline \\
Phenomenon & Ref. & WL & PI & Lifted PI  & PIL \\
\hline
Scalar implicature & \cite{grice1975} &  & x &x & x\\
Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguation & \cite{markman1988} &x  & & & x \\
Communicative equilibria & \cite{galantucci2005} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word learning model of \cite{frank2009}; PI refers to the pragmatic inference model of \cite{frank2012}; Lifted PI refers to the pragmatic inference model of \cite{bergen2012}, and PIL refers to the pragmatic inference and learning model described here.  }

@@

\subsection{Scalar implicature}

Many sets of words in natural language form scales in which each term refers to successively stronger restrictions on the situations compatible with the term. ``Some'' and ``all'' form a scale of this type. While ``I ate some of the cookies'' is compatible with the followup ``in fact, I ate {\it all} of the cookies,'' the reverse is not true. ``Might'' and ``must'' are another example, as are ``OK,'' ``good,'' and ``excellent.'' All of these scales allow for {\it scalar implicatures} \cite{grice1975}: if a speaker uses a weaker term, a listener is licensed to infer that the stronger term was not true in this situation. So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs.

The recursive pragmatic reasoning portions of our model capture set of findings on scalar implicature effectively. Consider a representation of scalar implicature as a matrix 
$\left[
    \begin{array}{cc}
      1 & 1 \\
      0 & 1
    \end{array} 
  \right]$,
where rows show the terms ``some'' and ``all'' and columns indicate worlds where {\sc some} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case. Assuming uniform costs for the two words and uniform priors, each recursion of listener modeling speaker reduces the probability that ``some'' will be used to refer to {\sc all}. Using the fixed parameters employed in our other simulations results in speaker use probabilities of 
$\left[
    \begin{array}{cc}
      1.00 & 0.00 \\
      0.00 & 1.00
    \end{array} 
  \right]$
and listener interpretation probabilities of 
$\left[
    \begin{array}{cc}
      .86 & .14 \\
      .03 & .97
    \end{array} 
  \right]$.

@@

Such an implicature again arises naturally from our model. We create a situation in which expressions have differential costs (e.g. $[.5 1.0]$) and referents have differential probabilities (e.g. $[.8 .2]$). Speakers in such a situation choose simple words to refer to common referents, e.g. 
$\left[
    \begin{array}{cc}
      .88 & .54 \\
      .12 & .46
    \end{array} 
  \right]$.\footnote{Rows indicate words and columns indicate referents. For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.} Similarly, listeners interpret ``cheap'' and ``expensive'' words to refer to common and rare objects respectively:
$\left[
    \begin{array}{cc}
      .77 & .23 \\
      .65 & .35
    \end{array} 
  \right]$.

@@

A recent line of work in language acquisition has 
% \begin{figure}[t]
%   \centering
%   \includegraphics[width=0.20\textwidth]{figures/cross-sit.pdf}
%   \caption{Cross-situational simulation}
%   \label{fig:cross-sit}
% \end{figure}

@@

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.20\textwidth]{figures/ME-1dax.pdf}
  \includegraphics[width=0.20\textwidth]{figures/ME-flat-10dog-10dax.pdf}
  \includegraphics[width=0.20\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
  \includegraphics[width=0.20\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
  \caption{Mutual exclusivity simulation}
  \label{fig:mutual-exclusivity}
\end{figure}

@@

>>>>

\documentclass{article} % For LaTeX2e
\usepackage{nips13submit_e,times}
\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[raggedright]{sidecap}

@@

\section{Experiments: Pragmatic and word learning phenomena}

@@

\begin{table}[t]
\label{tab:results}
\begin{center}
\begin{tabular}{lccccc}
\hline \\
Phenomenon & Ref. & WL & PI & Lifted PI  & PIL \\
\hline
Specificity implicature & \cite{grice1975} &  & x &x & x\\
Horn implicature & \cite{horn1984} & &  & x& x\\
Cross-situational learning & \cite{smith2008} & x& & & x \\
Disambiguation effect & \cite{markman1988} &x  & & & x \\
Disambiguation without mapping & \cite{horst2008} &  & & & x \\
Emergence of communicative equilibria & \cite{galantucci2005} & & & & x \\
Lexicalization of Horn implicature & \cite{horn1984} & & & & x \\
Lexicalization of specificity implicature & \cite{levinson2000} & & & & x \\
\hline
\end{tabular}
\end{center}
\caption{Empirical results and references. WL refers to the word learning model of \cite{frank2009}; PI refers to the pragmatic inference model of \cite{frank2012}; Lifted PI refers to the pragmatic inference model of \cite{bergen2012}, and PIL refers to the pragmatic inference and learning model described here.  }

@@

\subsection{Specificity implicature}

Many sets of words in natural language form scales in which each term refers to successively stronger restrictions on the situations compatible with the term. ``Some'' and ``all'' form a scale of this type. While ``I ate some of the cookies'' is compatible with the followup ``in fact, I ate {\it all} of the cookies,'' the reverse is not true. ``Might'' and ``must'' are another example, as are ``OK,'' ``good,'' and ``excellent.'' All of these scales allow for {\it scalar implicatures} \cite{grice1975}: if a speaker uses a weaker term, a listener is licensed to infer that the stronger term was not true in this situation. So although ``I ate some of the cookies'' could in principle be compatible with eating all of them, the listener is lead to believe that ``some but not all'' is the likely state of affairs. We use the term {\it specificity implicature} here (following \cite{bergen2012}) to denote that we are agnostic about whether such implicatures are conventionalized or whether they use a contextually-defined scale.

The recursive pragmatic reasoning portions of our model capture findings on specificity implicature. Consider a representation of scalar implicature as a matrix 
$\left[
    \begin{array}{cc}
      1 & 1 \\
      0 & 1
    \end{array} 
  \right]$,
where rows show the terms ``some'' and ``all'' and columns indicate worlds where {\sc some} and {\sc all} are true, respectively, capturing the regularity that ``some'' can be said of situations where {\sc all} is actually the case. Assuming uniform costs for the two words and uniform priors, each recursion of listener modeling speaker reduces the probability that ``some'' will be used to refer to {\sc all}. Using the fixed parameters employed in our other simulations results in speaker use probabilities of 
$\left[
    \begin{array}{cc}
      1.00 & 0.00 \\
      0.00 & 1.00
    \end{array} 
  \right]$
and listener interpretation probabilities of 
$\left[
    \begin{array}{cc}
      .86 & .14 \\
      .03 & .97
    \end{array} 
  \right]$.

@@

Such an implicature again arises naturally from our model. We create a situation in which expressions have differential costs (e.g. $[.5 1.0]$) and referents have differential probabilities (e.g. $[.8 .2]$). At the settings used throughout the manuscript, speakers in such a situation choose simple words to refer to common referents at different rates than for uncommon references. This situation breaks the symmetry though does not result in as dramatic a diagonal as in the specificity implicature, e.g. for the speaker, \footnote{Rows indicate words and columns indicate referents. For speakers, columns are normalized to indicate a distribution over words to be produced given referents, and for listeners rows are normalized to indicate a distribution on referents given words heard.}
$\left[
    \begin{array}{cc}
      .88 & .54 \\
      .12 & .46
    \end{array} 
  \right]$ 
and for the listener,
$\left[
    \begin{array}{cc}
      .77 & .23 \\
      .65 & .35
    \end{array} 
  \right]$. See Section \ref{sec:hornlearn} for the consequences of repeating this implicature until it is lexicalized.

@@

Even very young children can learn words from individually ambiguous situations, e.g. where multiple words are spoken and multiple objects are present \cite{smith2008}. While no individual situation gives away the mapping, the pattern of co-occurrences between words and objects allows for learning. Our current model reproduces this pattern of results, converging to correct lexical mappings from ambiguous evidence. A representative simulation is shown in Fig. \ref{fig:cross-sit}.

\begin{SCfigure}[2.5]
  \includegraphics[width=0.2\textwidth]{figures/cross-sit-mod.pdf}
  \caption{Results from a simulation of cross-situational word learning. The posterior distribution over the learner's lexicon (diagonal entries are correct for word/object pairs 1 -- 3; word D is a distractor). Training data were situations including two of the three objects and a single word, either the distractor or a word matching one of the two objects.}
  \label{fig:cross-sit}
\end{SCfigure}

@@

\begin{SCfigure}[1]
  \centering
  % \includegraphics[width=0.24\textwidth]{figures/ME-1dax.pdf}
  % \includegraphics[width=0.24\textwidth]{figures/ME-flat-10dog-10dax.pdf}
  % \includegraphics[width=0.24\textwidth]{figures/ME-antisparse-10dog-10dax.pdf}
  % \includegraphics[width=0.24\textwidth]{figures/ME-sparse-10dog-10dax.pdf}
\includegraphics[width=.6\textwidth]{figures/ME-composite-small.pdf}
  \caption{Results for simulations of the ``disambiguation'' phenomenon. A novel word is used in a situation with a novel object and a familiar object. Each plot shows the joint posterior probability that the familiar word refers to the familiar object (horizontal) and that the novel word refers to the familiar object (vertical). Left plot shows the posterior after a single situation; right plot shows multiple situations. 
%The right middle plot shows the posterior after a single situation given either a prior that makes it likely that there are words referring to both the novel and familiar objects (perhaps because the two objects are physically similar; the right plot shows a situation where the prior makes it unlikely that a word refers to both (objects dissimilar). 
}
  \label{fig:me}
\end{SCfigure}

Children, when presented with both a novel and a familiar object (e.g. an eggbeater and a ball), will treat a novel label (e.g. ``dax'') as referring to the novel object, for example by supplying the eggbeater when asked to ``give me the dax'' \cite{markman1988}. This phenomenon is sometimes referred to as ``mutual exclusivity'' or even ``Fast mapping,'' and many distinct theoretical accounts have been proposed in the psychological literature. Simple probabilistic word learning models can produce a similar pattern of findings \cite{frank2009}, often due to the kinds of ``explaining away'' type phenomena that are observed in such models. But all such models assume that learners in fact retain the mapping between word and object that is demonstrated in the experimental situation; this observation is contradicted, however, by evidence that children often do not retain the mappings that are demonstrated by their inferences in the moment \cite{horst2008}.

Our model provides an intriguing possible explanation of this finding: given a single disambiguation situation, the model gives a substantial probability (e.g. 75\%) that the speaker is referring to the novel object. Nevertheless, integrating across lexicons reveals that this probability is not due to learning that the novel word refers to this object (as shown in Fig. \ref{fig:me}, left): instead, it comes in part from a specificity implicature because the listener knows that the familiar word \emph{does not} refer to the novel object--hence the novel word (even if it refers to the familiar object as well) is the best way to refer to the novel object. Nevertheless, on repeated exposure to the same novel word, novel object situation, the model does lexicalize this inference (Fig. \ref{fig:me}, right).

@@

@@

\newpage
